---
title: "Retrospective Evaluation of Soccer Players (REvSoc)"
author: "Seshadhri S"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---
Github Credentials:
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


# 2024-25 season
## Goalkeeper Rating System: Methodology & Validation
In this analysis, I developed a multidimensional goalkeeper rating model designed to evaluate both conventional and modern aspects of goalkeeping performance. The model incorporates traditional shot-stopping metrics alongside advanced measures of distribution, sweeping actions, and cross management to reflect the evolving demands of the position.

### Method Overview
The rating was divided into two core components:

Conventional Keeper Rating (CKR) — incorporating metrics like Saves, Goals Against (GA), Save Percentage, and Post-Shot Expected Goals (PSxG) overperformance.

Sweeper Keeper Rating (SKR) — capturing distribution accuracy, launch rates, defensive actions outside the box, average distance of defensive actions, and cross-stopping percentages.

Both components were normalized relative to their averages and combined to produce a Total Goalkeeper Rating.

### Validation Process

To ensure the model was both credible and informative, I validated the Total Goalkeeper Rating against three independent metrics:

1. Goals Against (GA)

2. PSxG-GA per 90 (shot-stopping over expected)

3. Save Percentage


```{r}

library(worldfootballR)
library(purrr)
library(dplyr)
library(Boruta)
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(ggrepel)

```

## Load in FBRef Data for the Big 5 Leagues

```{r}
big5_player_standard <- fb_big5_advanced_season_stats(season_end_year= 2024, stat_type= "standard", team_or_player= "player")
big5_player_shooting <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "shooting", team_or_player= "player")
big5_player_passing <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "passing", team_or_player= "player")
big5_player_passing_types <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "passing_types", team_or_player= "player")
big5_player_gca <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "gca", team_or_player= "player")
big5_player_defense <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "defense", team_or_player= "player")
big5_player_possession <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "possession", team_or_player= "player")
big5_player_playing_time <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "playing_time", team_or_player= "player")
big5_player_misc <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "misc", team_or_player= "player")
big5_player_keepers <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "keepers", team_or_player= "player")
big5_player_keepers_adv <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "keepers_adv", team_or_player= "player")

prem_league_table <- fb_season_team_stats(country = "ENG", gender = "M", season_end_year = "2025", tier = "1st", stat_type = "league_table") %>% arrange(desc(Pts))
seriea_league_table <- fb_season_team_stats(country = "ITA", gender = "M", season_end_year = "2025", tier = "1st", stat_type = "league_table") %>% arrange(desc(Pts))
ligue1_league_table <- fb_season_team_stats(country = "FRA", gender = "M", season_end_year = "2025", tier = "1st", stat_type = "league_table") %>% arrange(desc(Pts))
laliga_league_table <- fb_season_team_stats(country = "ESP", gender = "M", season_end_year = "2025", tier = "1st", stat_type = "league_table") %>% arrange(desc(Pts))
bundesliga_league_table <- fb_season_team_stats(country = "GER", gender = "M", season_end_year = "2025", tier = "1st", stat_type = "league_table") %>% arrange(desc(Pts))

```

## Goalkeeper Ratings

```{r}
##------Goalkeeper----
#>Have to divide this into Conventional Keeper Rating (CKR) and Sweeper Keeper Rating (SKR)
#>Take the keepers away from the playing times dataframe

gk_playing_time <- big5_player_playing_time %>% filter(Pos == "GK") %>% filter(MP_Playing.Time != 0)

df_list <- list(gk_playing_time, big5_player_keepers, big5_player_keepers_adv)
common_columns <- Reduce(intersect, lapply(df_list, colnames))

merged_df <- inner_join(big5_player_keepers, big5_player_keepers_adv, gk_playing_time, by = c("Season_End_Year", "Squad", "Comp", "Player", "Nation", "Pos", "Age", "Born", "Url"))

gk_above_900 <- merged_df %>% filter(Min_Playing >= 900)

keepers_numeric <-gk_above_900 %>% select(where(is.numeric))

# Optional: remove NAs
keepers_numeric <- keepers_numeric[, -1]
keepers_numeric <- na.omit(keepers_numeric)

# Scale the data (important for PCA since units differ)
keepers_scaled <- scale(keepers_numeric)
keepers_scaled <- keepers_scaled[ ,-1]
# Run PCA
keepers_pca <- PCA(keepers_scaled, graph = FALSE)

# View eigenvalues (variance explained by each component)
keepers_pca$eig

# Extract variance explained by Dim.1 and Dim.2
dim1_variance <- keepers_pca$eig[1, "percentage of variance"]
dim2_variance <- keepers_pca$eig[2, "percentage of variance"]

# Normalize so they sum to 1 (as relative weights)
total_variance <- dim1_variance + dim2_variance
dim1_weight <- dim1_variance / total_variance
dim2_weight <- dim2_variance / total_variance

print(paste("Dim.1 weight:", round(dim1_weight, 3)))
print(paste("Dim.2 weight:", round(dim2_weight, 3)))



# Plot the variance explained
fviz_eig(keepers_pca)

# Biplot (players and variables on first two PCs)
fviz_pca_biplot(keepers_pca, repel = TRUE)

# Contributions of variables to PC1 and PC2
fviz_pca_var(keepers_pca, 
             col.var = "contrib",
             labelsize = 2, repel = TRUE) 

variable_contrib <- data.frame(keepers_pca$var$contrib)
variable_contrib <- variable_contrib %>%
  mutate(Total_Contrib = Dim.1 + Dim.2)

# Scale them to sum to 1 (to act as relative weights)
variable_contrib <- variable_contrib %>%
  mutate(Weight = Total_Contrib / sum(Total_Contrib))

# 75th percentile threshold
threshold_75 <- quantile(variable_contrib$Total_Contrib, 0.75)
print(threshold_75)

# Sorted plot
sorted_scores <- sort(variable_contrib$Total_Contrib, decreasing = TRUE)
plot(sorted_scores, type = "l", col = "blue", main = "PCA Score Distribution",
     ylab = "PCA Score (Dim1 + Dim2)", xlab = "Ranked Goalkeepers")


# Find the elbow point
elbow <- inflection::findiplist(1:length(sorted_scores), sorted_scores, 0)
print(elbow)

mean_score <- mean(variable_contrib$Total_Contrib)
sd_score <- sd(variable_contrib$Total_Contrib)

threshold_sd <- mean_score + sd_score
print(threshold_sd)

plot(density(variable_contrib$Total_Contrib), main = "Density of PCA Scores")
abline(v = mean_score, col = "red", lty = 2)

```

The PCA variable contribution plot shows a clear differentiation between shot stopping and sweeper keeper metrics

1. OPA_per_90_Sweeper and AvgDist_Sweeper are almost overlapping and long - These two are very closely related and both contribute heavily to PC1.
TL;DR: your sweeper keeper stats move together and drive a big chunk of the differences between keepers.

2. PSxG+_per__minus__Expected (the expected goals prevented stat) points in a different direction and is quite long too - Strongly influential but not correlated with the sweeper actions. It’s carrying weight independently, probably defining a classic shot-stopper profile.

3. Cmp_percent_Launched and Stp_percent_Crosses have shorter vectors, contributing less to the variance on these first two components.
Could mean they’re either:

- Less variable between your keepers

- Or important on other PCs (PC3, PC4…)
```{r}

CKR <- with(gk_above_900,
  (Min_Playing * 
   Saves * variable_contrib["Saves", "Weight"] *
   PSxG_Expected * variable_contrib["PSxG_Expected", "Weight"]) /
  (GA * variable_contrib["GA", "Weight"] * GA * SoTA * variable_contrib["SoTA", "Weight"])
)

SKR <- with(gk_above_900,
  (
    (Cmp_Launched/Att_Launched) * variable_contrib["Cmp_Launched", "Weight"] +
    (Stp_Crosses/Opp_Crosses) * variable_contrib["Stp_Crosses", "Weight"] +
    (AvgLen_Passes/`Att (GK)_Passes`) * variable_contrib["AvgLen_Passes", "Weight"] +
    (AvgDist_Sweeper/`#OPA_Sweeper`) * variable_contrib["AvgDist_Sweeper", "Weight"] +
    (AvgLen_Goal/Att_Goal) * variable_contrib["AvgLen_Goal", "Weight"]
  ) / Min_Playing
)

avg_CKR <- mean(CKR, na.rm = TRUE)
avg_SKR <- mean(SKR, na.rm = TRUE)

CKR_Avg <- CKR / avg_CKR
SKR_Avg <- SKR / avg_SKR

Total_GK_Rating <- (CKR_Avg * dim1_weight) + (SKR_Avg * dim2_weight)

gk_ratings_final <- gk_above_900 %>% 
  mutate(
    CKR_Avg = CKR_Avg,
    SKR_Avg = SKR_Avg,
    Total_GK_Rating = Total_GK_Rating
  ) %>%
  arrange(desc(Total_GK_Rating))

gk_ratings_24starts <- gk_ratings_final %>% filter(Starts_Playing >=25) %>% select(Squad, Comp, Player, CKR_Avg, SKR_Avg, Total_GK_Rating)
print(head(gk_ratings_24starts))

#Correlation work between GA and GKR

cor(gk_ratings_final$Total_GK_Rating, gk_ratings_final$GA, use = "complete.obs", method = "spearman")
cor(gk_ratings_final$Total_GK_Rating, gk_ratings_final$`PSxG+_per__minus__Expected`, use = "complete.obs", method = "spearman")
cor(gk_ratings_final$Total_GK_Rating, (gk_ratings_final$PSxG_Expected - gk_ratings_final$GA) / (gk_ratings_final$Min_Playing / 90), method = "spearman")
cor(gk_ratings_final$Total_GK_Rating, gk_ratings_final$Save_percent, use = "complete.obs", method = "spearman")


gk_ratings_league_avg <- gk_ratings_final %>% group_by(Comp) %>% summarise(Avg_GKR = mean(Total_GK_Rating)) 
print(gk_ratings_league_avg)
gk_psxg_league_avg <- gk_ratings_final %>% group_by(Comp) %>% summarise(Avg_PSxG = mean(PSxG_Expected))
print(gk_psxg_league_avg)
cor(gk_ratings_league_avg$Avg_GKR, gk_psxg_league_avg$Avg_PSxG, method = "spearman")

#Split them into Each League

pl_gk_ratings <- gk_ratings_final %>% filter(Comp == "Premier League")
bundesliga_gk_ratings <- gk_ratings_final %>% filter(Comp == "Bundesliga")
seriea_gk_ratings <- gk_ratings_final %>% filter(Comp == "Serie A")
ligue1_gk_ratings <- gk_ratings_final %>% filter(Comp == "Ligue 1")
laliga_gk_ratings <- gk_ratings_final %>% filter(Comp == "La Liga")

#Awards for each league
pl_gk_ratings_awards <- gk_ratings_24starts %>% filter(Comp == "Premier League")
bundesliga_gk_ratings_awards <- gk_ratings_24starts %>% filter(Comp == "Bundesliga")
seriea_gk_ratings_awards <- gk_ratings_24starts %>% filter(Comp == "Serie A")
ligue1_gk_ratings_awards <- gk_ratings_24starts %>% filter(Comp == "Ligue 1")
laliga_gk_ratings_awards <- gk_ratings_24starts %>% filter(Comp == "La Liga")
```

### Interpretation

These correlations confirm that the model is strongly aligned with key outcome metrics while maintaining balance:

High correlation with Goals Against indicates outcome relevance.

Moderate correlations with PSxG-GA per 90 and Save Percentage confirm the model’s shot-stopping validity without overfitting to a single dimension.

The moderate strength of these relationships reflects the rating’s inclusion of modern, non-shot-stopping actions, enhancing its relevance to contemporary goalkeeper profiles.

### Conclusion

This multidimensional rating offers a more comprehensive view of goalkeeper performance than conventional single-metric models. It captures the hybrid nature of the modern goalkeeper — one expected to contribute both in traditional shot-stopping and proactive defensive play — and stands up well to statistical validation against key performance outcomes.

## Outfield Player Ratings
### Defensive Player Ratings

```{r}
#>DEFENSIVE PLAYER RATING

outfield_playing_time <- big5_player_playing_time %>% filter(Pos != "GK") %>% filter(Min_Playing.Time >=900)
outfield_df_list <- list(outfield_playing_time, big5_player_defense, big5_player_gca, big5_player_misc, 
                         big5_player_passing, big5_player_passing_types, big5_player_possession,
                         big5_player_shooting, big5_player_standard)
common_columns_outfield <- Reduce(intersect, lapply(outfield_df_list, colnames))
print(common_columns_outfield)

outfield_data <- outfield_df_list %>%
  reduce(left_join, by = c("Season_End_Year", "Squad", "Comp", "Player", "Nation", "Pos", "Age", "Born", "Url"))
outfield_data <- outfield_data %>%
  select(-Url)

outfield_defensive_data <- outfield_data %>% select(Season_End_Year, Squad, Comp, Player, Nation, Pos, Age, Min_Playing.Time, Tkl_Tackles, TklW_Tackles, Tkl_Challenges, Att_Challenges, Blocks_Blocks, Int.x, Clr, Err, CrdY.y, CrdR.y, Recov, Fls, Fld, Recov, Won_Aerial, Lost_Aerial)

outfield_above_900 <- outfield_defensive_data %>% filter(Min_Playing.Time >= 900)

outfield_numeric <-outfield_above_900 %>% select(where(is.numeric))

# Optional: remove NAs
outfield_numeric <- outfield_numeric[, -1]
outfield_numeric <- na.omit(outfield_numeric)

# Scale the data (important for PCA since units differ)
outfield_scaled <- scale(outfield_numeric)
outfield_scaled <- outfield_scaled[ ,-1]
# Run PCA
outfield_pca <- PCA(outfield_scaled, graph = FALSE)

# View eigenvalues (variance explained by each component)
outfield_pca$eig

# Extract variance explained by Dim.1 and Dim.2
dim1_variance <- outfield_pca$eig[1, "percentage of variance"]
dim2_variance <- outfield_pca$eig[2, "percentage of variance"]

# Normalize so they sum to 1 (as relative weights)
total_variance <- dim1_variance + dim2_variance
dim1_weight <- dim1_variance / total_variance
dim2_weight <- dim2_variance / total_variance

print(paste("Dim.1 weight:", round(dim1_weight, 3)))
print(paste("Dim.2 weight:", round(dim2_weight, 3)))



# Plot the variance explained
fviz_eig(outfield_pca)

# Biplot (players and variables on first two PCs)
fviz_pca_biplot(outfield_pca, repel = TRUE)

# Contributions of variables to PC1 and PC2
fviz_pca_var(outfield_pca, 
             col.var = "contrib",
             labelsize = 2, repel = TRUE) 

variable_contrib <- data.frame(outfield_pca$var$contrib)
variable_contrib <- variable_contrib %>%
  mutate(Total_Contrib = Dim.1 + Dim.2)

# Scale them to sum to 1 (to act as relative weights)
variable_contrib <- variable_contrib %>%
  mutate(Weight = Total_Contrib / sum(Total_Contrib))

# 75th percentile threshold
threshold_75 <- quantile(variable_contrib$Total_Contrib, 0.75)
print(threshold_75)

# Sorted plot
sorted_scores <- sort(variable_contrib$Total_Contrib, decreasing = TRUE)
plot(sorted_scores, type = "l", col = "blue", main = "PCA Score Distribution",
     ylab = "PCA Score (Dim1 + Dim2)", xlab = "Ranked Goalkeepers")


# Find the elbow point
elbow <- inflection::findiplist(1:length(sorted_scores), sorted_scores, 0)
print(elbow)

mean_score <- mean(variable_contrib$Total_Contrib)
sd_score <- sd(variable_contrib$Total_Contrib)

threshold_sd <- mean_score + sd_score
print(threshold_sd)

plot(density(variable_contrib$Total_Contrib), main = "Density of PCA Scores")
abline(v = mean_score, col = "red", lty = 2)
#########
```


```{r}

outfield_dpr <- outfield_data %>% mutate(Defensive_Player_Rtg = ((Tkl_Tackles* variable_contrib["Tkl_Tackles", "Weight"]/TklW_Tackles* variable_contrib["TklW_Tackles", "Weight"])+(Tkl_Challenges* variable_contrib["Tkl_Challenges", "Weight"]/Att_Challenges* variable_contrib["Att_Challenges", "Weight"])+Blocks_Blocks* variable_contrib["Blocks_Blocks", "Weight"]+Int.x* variable_contrib["Int.x", "Weight"]+ Clr* variable_contrib["Clr", "Weight"]-Err*variable_contrib["Err", "Weight"]-CrdY.y* variable_contrib["CrdY.y", "Weight"]-(2*(CrdR.y* variable_contrib["CrdR.y", "Weight"]))-(Fls*variable_contrib["Fls", "Weight"]/Fld*variable_contrib["Fld", "Weight"]) + Recov* variable_contrib["Recov", "Weight"] + (Won_Aerial* variable_contrib["Won_Aerial", "Weight"]/Lost_Aerial* variable_contrib["Lost_Aerial", "Weight"])))


outfield_dpr_filtered <- outfield_dpr %>%
  filter(is.finite(Defensive_Player_Rtg), !is.na(Defensive_Player_Rtg))


#> Get the Average possession for every team in the top 5 leagues (will be useful for other metrics too)
#> For defenders, divide DPR by (100 - Avg. Poss.)
#> First things first, let's extract the possession data at team level

big5_team_possession <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "possession", team_or_player= "team")
dplyr::glimpse(big5_team_possession)
big5_team_possession_opponent <- big5_team_possession %>% filter(Team_or_Opponent == "opponent") %>% select(Squad, Poss)
big5_team_possession_team <- big5_team_possession %>% filter(Team_or_Opponent == "team") %>% select(Squad, Poss)

outfield_dpr_filtered_possopp <- outfield_dpr_filtered %>% left_join(big5_team_possession_opponent, by = "Squad", relationship = "many-to-many") %>% rename(Poss_Opp = Poss)
outfield_dpr_filtered_possopp_possteam <- outfield_dpr_filtered_possopp %>% left_join(big5_team_possession_team, by = "Squad", relationship = "many-to-many") %>% rename(Poss_Team = Poss)

outfield_DPR <- outfield_dpr_filtered_possopp_possteam %>% mutate(Poss.Adj.DPR = Defensive_Player_Rtg*100/Poss_Opp) %>% arrange(desc(Poss.Adj.DPR))
avg_dpr <- mean(outfield_DPR$Poss.Adj.DPR)
outfield_DPR <- outfield_DPR %>% mutate(Poss.Adj.DPR = (Poss.Adj.DPR/avg_dpr)) %>% arrange(desc(Poss.Adj.DPR))


pl_dpr <- outfield_DPR %>% filter(Comp == "Premier League")
ligue1_dpr <- outfield_DPR %>% filter(Comp == "Ligue 1")
seriea_dpr <- outfield_DPR %>% filter(Comp == "Serie A")
bundesliga_dpr <- outfield_DPR %>% filter(Comp == "Bundesliga")
laliga_dpr <- outfield_DPR %>% filter(Comp == "La Liga")

#> Outfield_DPR is what you have to carry forward because you cannot have players ineligible for any category when it comes to final team ratings

```

### Offensive Creation Rating

```{r}
#>OFFENSIVE CREATION RATING

outfield_ocr_data <- outfield_data %>% select(Season_End_Year, Squad, Comp, Player, Nation, Pos, Age, Min_Playing.Time, Ast.x, xA_Expected, SCA_SCA, GCA_GCA, KP, Final_Third, PPA, Att_Total, CrsPA, Crs, PrgDist_Total, TotDist_Total, `Att 3rd_Touches`, `Att Pen_Touches`, Touches_Touches, Succ_Take, Att_Take, PrgDist_Carries, TotDist_Carries, Final_Third_Carries, PrgC_Carries, CPA_Carries, Carries_Carries, PrgR_Receiving, Rec_Receiving, Mis_Carries, Dis_Carries, Carries_Carries, PKwon)

colnames(outfield_ocr_data)[21] <- "Att_3rd_Touches"
colnames(outfield_ocr_data)[22] <- "Att_Pen_Touches"

outfield_above_900 <- outfield_ocr_data%>% filter(Min_Playing.Time >= 900)
outfield_above_900_complete <- na.omit(outfield_above_900)

outfield_numeric <-outfield_above_900 %>% select(where(is.numeric))

# Optional: remove NAs
outfield_numeric <- outfield_numeric[, -1]
outfield_numeric <- na.omit(outfield_numeric)

# Scale the data (important for PCA since units differ)
outfield_scaled <- scale(outfield_numeric)
outfield_scaled <- outfield_scaled[ ,-1]
# Run PCA
outfield_pca <- PCA(outfield_scaled, graph = FALSE)

# View eigenvalues (variance explained by each component)
outfield_pca$eig

# Extract variance explained by Dim.1 and Dim.2
dim1_variance <- outfield_pca$eig[1, "percentage of variance"]
dim2_variance <- outfield_pca$eig[2, "percentage of variance"]

# Normalize so they sum to 1 (as relative weights)
total_variance <- dim1_variance + dim2_variance
dim1_weight <- dim1_variance / total_variance
dim2_weight <- dim2_variance / total_variance

print(paste("Dim.1 weight:", round(dim1_weight, 3)))
print(paste("Dim.2 weight:", round(dim2_weight, 3)))



# Plot the variance explained
fviz_eig(outfield_pca)

# Biplot (players and variables on first two PCs)
fviz_pca_biplot(outfield_pca, repel = TRUE)

# Contributions of variables to PC1 and PC2
fviz_pca_var(outfield_pca, 
             col.var = "contrib",
             labelsize = 2, repel = TRUE) 

variable_contrib <- data.frame(outfield_pca$var$contrib)
variable_contrib <- variable_contrib %>%
  mutate(Total_Contrib = Dim.1 + Dim.2)

# Scale them to sum to 1 (to act as relative weights)
variable_contrib <- variable_contrib %>%
  mutate(Weight = Total_Contrib / sum(Total_Contrib))

# 75th percentile threshold
threshold_75 <- quantile(variable_contrib$Total_Contrib, 0.75)
print(threshold_75)

# Sorted plot
sorted_scores <- sort(variable_contrib$Total_Contrib, decreasing = TRUE)
plot(sorted_scores, type = "l", col = "blue", main = "PCA Score Distribution",
     ylab = "PCA Score (Dim1 + Dim2)", xlab = "Ranked Goalkeepers")


# Find the elbow point
elbow <- inflection::findiplist(1:length(sorted_scores), sorted_scores, 0)
print(elbow)

mean_score <- mean(variable_contrib$Total_Contrib)
sd_score <- sd(variable_contrib$Total_Contrib)

threshold_sd <- mean_score + sd_score
print(threshold_sd)

plot(density(variable_contrib$Total_Contrib), main = "Density of PCA Scores")
abline(v = mean_score, col = "red", lty = 2)

###########
```

```{r}
#> OFFENSIVE CREATION RATING

outfield_DPR_ocr <- outfield_above_900 %>% mutate(Off_Creation_Rtg = (((Ast.x *variable_contrib["Ast.x", "Weight"])/(xA_Expected* variable_contrib["xA_Expected", "Weight"]))+ ((SCA_SCA* variable_contrib["SCA_SCA", "Weight"])) + ((GCA_GCA* variable_contrib["GCA_GCA", "Weight"])) +(((KP* variable_contrib["KP", "Weight"]) + (Final_Third * variable_contrib["Final_Third", "Weight"]) + (PPA * variable_contrib["PPA", "Weight"]))/(Att_Total* variable_contrib["Att_Total", "Weight"]))+((CrsPA* variable_contrib["CrsPA", "Weight"])/(Crs* variable_contrib["Crs", "Weight"]))+ ((PrgDist_Total* variable_contrib["PrgDist_Total", "Weight"])/(TotDist_Total* variable_contrib["TotDist_Total", "Weight"]))+ (((Att_3rd_Touches* variable_contrib["Att_3rd_Touches", "Weight"]) + (Att_Pen_Touches* variable_contrib["Att_Pen_Touches", "Weight"]))/(Touches_Touches* variable_contrib["Touches_Touches", "Weight"]))+ ((Succ_Take* variable_contrib["Succ_Take", "Weight"])/(Att_Take* variable_contrib["Att_Take", "Weight"]))+((PrgDist_Carries* variable_contrib["PrgDist_Carries", "Weight"])/(TotDist_Carries* variable_contrib["TotDist_Carries", "Weight"]))+(((Final_Third_Carries* variable_contrib["Final_Third_Carries", "Weight"]) + (PrgC_Carries* variable_contrib["PrgC_Carries", "Weight"]) + (CPA_Carries* variable_contrib["CPA_Carries", "Weight"]))/(Carries_Carries* variable_contrib["Carries_Carries", "Weight"]))+ ((PrgR_Receiving* variable_contrib["PrgR_Receiving", "Weight"])/(Rec_Receiving* variable_contrib["Rec_Receiving", "Weight"])) - (((Mis_Carries* variable_contrib["Mis_Carries", "Weight"])+(Dis_Carries* variable_contrib["Dis_Carries", "Weight"]))/(Carries_Carries* variable_contrib["Carries_Carries", "Weight"])) + (PKwon* variable_contrib["PKwon", "Weight"]))) %>% arrange(desc(Off_Creation_Rtg))

setdiff(c("Ast.x", "xA_Expected", "SCA_SCA", "GCA_GCA", "KP", "Final_Third", "PPA", "Att_Total", "CrsPA", "Crs", "PrgDist_Total", "TotDist_Total", "Att_3rd_Touches", "Att_Pen_Touches", "Touches_Touches", "Succ_Take", "Att_Take", "TotDist_Carries", "Final_Third_Carries", "PrgC_Carries", "CPA_Carries","Carries_Carries", "PrgR_Receiving","Rec_Receiving", "Mis_Carries", "Dis_Carries", "PKwon"), rownames(variable_contrib))

big5_team_possession <- fb_big5_advanced_season_stats(season_end_year= 2025, stat_type= "possession", team_or_player= "team")
dplyr::glimpse(big5_team_possession)
big5_team_possession_opponent <- big5_team_possession %>% filter(Team_or_Opponent == "opponent") %>% select(Squad, Poss)
big5_team_possession_team <- big5_team_possession %>% filter(Team_or_Opponent == "team") %>% select(Squad, Poss)

outfield_DPR_ocr <- outfield_DPR_ocr %>% left_join(big5_team_possession_opponent, by = "Squad", relationship = "many-to-many") %>% rename(Poss_Opp = Poss)
outfield_DPR_ocr <- outfield_DPR_ocr %>% left_join(big5_team_possession_team, by = "Squad", relationship = "many-to-many") %>% rename(Poss_Team = Poss)

outfield_DPR_ocr_possadj <- outfield_DPR_ocr %>% mutate(Poss.adj.OCR = (Off_Creation_Rtg/Poss_Team)*100) %>% arrange(desc(Poss.adj.OCR))
outfield_DPR_ocr_possadj <- outfield_DPR_ocr_possadj %>% filter(Poss.adj.OCR != "NaN")
avg_ocr <- mean(outfield_DPR_ocr_possadj$Poss.adj.OCR)
outfield_DPR_OCR <- outfield_DPR_ocr_possadj %>% mutate(Poss.adj.OCR = Poss.adj.OCR/avg_ocr) %>% arrange(desc(Poss.adj.OCR))

outfield_DPR_OCR_5ast <- outfield_DPR_OCR %>% filter(Ast.x >= 5)

```

### Goalscorer Rating

```{r}
#>GOALSCORER RATING

outfield_gsr_data <- outfield_data %>%
  mutate(
    Goals_per_xG = ifelse(xG_Expected.x > 0, Gls_Standard / xG_Expected.x, 0),
    G_per_Shot = ifelse(Sh_Standard > 0, Gls_Standard / Sh_Standard, 0),
    xG_per_Shot = ifelse(Sh_Standard > 0, xG_Expected.x / Sh_Standard, 0))


outfield_gsr_data <- outfield_gsr_data %>% mutate(Gls_per_xG = Gls/xG_Expected.x) %>% select(Season_End_Year, Squad, Comp, Player, Nation, Pos, Age, Min_Playing.Time, Gls, xG_Expected.x, SoT_percent_Standard, Sh_Standard, Gls_Per, xG_Per, PrgR_Receiving, Gls_per_xG, Goals_per_xG, G_per_Shot, xG_per_Shot) %>% filter(!is.nan(Gls_per_xG))

colnames(outfield_gsr_data)[10] <- "xG_Expected"

outfield_above_900 <- outfield_gsr_data%>% filter(Min_Playing.Time >= 900)
outfield_above_900_complete <- na.omit(outfield_above_900)

outfield_numeric <-outfield_above_900 %>% select(where(is.numeric))

# Optional: remove NAs
outfield_numeric <- outfield_numeric[, -1]
outfield_numeric <- na.omit(outfield_numeric)

# Scale the data (important for PCA since units differ)
outfield_scaled <- scale(outfield_numeric)
outfield_scaled <- outfield_scaled[ ,-1]
# Run PCA
outfield_pca <- PCA(outfield_scaled, graph = FALSE)

# View eigenvalues (variance explained by each component)
outfield_pca$eig

# Extract variance explained by Dim.1 and Dim.2
dim1_variance <- outfield_pca$eig[1, "percentage of variance"]
dim2_variance <- outfield_pca$eig[2, "percentage of variance"]

# Normalize so they sum to 1 (as relative weights)
total_variance <- dim1_variance + dim2_variance
dim1_weight <- dim1_variance / total_variance
dim2_weight <- dim2_variance / total_variance

print(paste("Dim.1 weight:", round(dim1_weight, 3)))
print(paste("Dim.2 weight:", round(dim2_weight, 3)))



# Plot the variance explained
fviz_eig(outfield_pca)

# Biplot (players and variables on first two PCs)
fviz_pca_biplot(outfield_pca, repel = TRUE)

# Contributions of variables to PC1 and PC2
fviz_pca_var(outfield_pca, 
             col.var = "contrib",
             labelsize = 2, repel = TRUE) 

variable_contrib <- data.frame(outfield_pca$var$contrib)
variable_contrib <- variable_contrib %>%
  mutate(Total_Contrib = Dim.1 + Dim.2)

# Scale them to sum to 1 (to act as relative weights)
variable_contrib <- variable_contrib %>%
  mutate(Weight = Total_Contrib / sum(Total_Contrib))

# 75th percentile threshold
threshold_75 <- quantile(variable_contrib$Total_Contrib, 0.75)
print(threshold_75)

# Sorted plot
sorted_scores <- sort(variable_contrib$Total_Contrib, decreasing = TRUE)
plot(sorted_scores, type = "l", col = "blue", main = "PCA Score Distribution",
     ylab = "PCA Score (Dim1 + Dim2)", xlab = "Ranked Goalkeepers")


# Find the elbow point
elbow <- inflection::findiplist(1:length(sorted_scores), sorted_scores, 0)
print(elbow)

mean_score <- mean(variable_contrib$Total_Contrib)
sd_score <- sd(variable_contrib$Total_Contrib)

threshold_sd <- mean_score + sd_score
print(threshold_sd)

plot(density(variable_contrib$Total_Contrib), main = "Density of PCA Scores")
abline(v = mean_score, col = "red", lty = 2)

###########

```

```{r}
#> GOALSCORER RATING
#Verify if the goals are per 90 or per shot - think they might be per 90
outfield_DPR_OCR_gsr <- outfield_gsr_data %>% filter(Min_Playing.Time > 900) %>% mutate(GoalScorer_Rtg = ((Gls * variable_contrib["Gls", "Weight"]) +
    (ifelse(xG_Expected > 0, Gls / xG_Expected, 0) * variable_contrib["Gls_per_xG", "Weight"]) +
    (SoT_percent_Standard * variable_contrib["SoT_percent_Standard", "Weight"]) +
    (ifelse(Sh_Standard > 0, Gls / Sh_Standard, 0) * variable_contrib["G_per_Shot", "Weight"]) +
    (xG_per_Shot * variable_contrib["xG_per_Shot", "Weight"]) +
    (PrgR_Receiving * variable_contrib["PrgR_Receiving", "Weight"])
) / Min_Playing.Time) %>% arrange(desc(GoalScorer_Rtg))


outfield_gsr_awards <- outfield_DPR_OCR_gsr %>% filter(Min_Playing.Time >= 2160)
outfield_GSR <- outfield_DPR_OCR_gsr #Save this object alone
```

#### Sanity check GSR with xG

```{r}
#Correlation work between GA and GKR

cor(outfield_GSR$GoalScorer_Rtg, outfield_GSR$Gls_Per, use = "complete.obs", method = "spearman")
cor(outfield_GSR$GoalScorer_Rtg, outfield_GSR$xG_Per, use = "complete.obs", method = "spearman")

summary_model_GSR <- lm(Gls_Per ~ GoalScorer_Rtg, data = outfield_GSR)
summary(summary_model_GSR)$r.squared

```

### Reassigning a few table names

```{r}
outfield_OCR <- outfield_DPR_OCR
```

## Overall Player Ratings
#### Take the four ratings dataframes and join them before doing clustering on them

```{r}
outfield_DPR_filtered <- outfield_DPR %>% select(Squad, Comp, Player, Pos, Poss.Adj.DPR, Min_Playing.Time)
outfield_OCR_filtered <- outfield_OCR %>% select(Squad, Comp, Player, Pos, Poss.adj.OCR, Min_Playing.Time)
outfield_GSR_filtered <- outfield_GSR %>% select(Squad, Comp, Player, Pos, GoalScorer_Rtg, Min_Playing.Time)

player_master <- full_join(outfield_DPR_filtered, outfield_OCR_filtered, by = "Player") %>% full_join(outfield_GSR_filtered, by = "Player") 

player_master <- player_master %>%
  mutate(
    DPR_std = (Poss.Adj.DPR - min(Poss.Adj.DPR, na.rm=TRUE)) / (max(Poss.Adj.DPR, na.rm=TRUE) - min(Poss.Adj.DPR, na.rm=TRUE)),
    OCR_std = (Poss.adj.OCR - min(Poss.adj.OCR, na.rm=TRUE)) / (max(Poss.adj.OCR, na.rm=TRUE) - min(Poss.adj.OCR, na.rm=TRUE)),
    GSR_std = (GoalScorer_Rtg - min(GoalScorer_Rtg, na.rm=TRUE)) / (max(GoalScorer_Rtg, na.rm=TRUE) - min(GoalScorer_Rtg, na.rm=TRUE))
  )

player_master <- player_master %>%  mutate(
    Squad= coalesce(Squad, Squad.x, Squad.y),
    Comp= coalesce(Comp, Comp.x, Comp.y),
    Pos = coalesce(Pos.x, Pos.y, Pos),
    Min_Playing_Time = coalesce(Min_Playing.Time.x, Min_Playing.Time.y, Min_Playing.Time))

player_master_final <- player_master %>% select(Player, Squad, Comp, Pos, Poss.Adj.DPR, Poss.adj.OCR, GoalScorer_Rtg, DPR_std, OCR_std, GSR_std, Min_Playing_Time)
```


### Cluster the players to look for any natural clusterings then run PCA on each cluster

```{r}
player_clustering <- player_master_final %>%
  select(DPR_std, OCR_std, GSR_std) %>%
  mutate(across(everything(), ~ ifelse(is.na(.), 0, .)))

wss <- function(k) {
  kmeans(player_clustering, k, nstart = 25)$tot.withinss
}

k.values <- 1:10
wss_values <- map_dbl(k.values, wss)

plot(k.values, wss_values, type = "b",
     pch = 19, frame = FALSE, 
     xlab = "Number of clusters K",
     ylab = "Total within-clusters sum of squares")

library(cluster)

sil_width <- function(k){
  pam(player_clustering, k)$silinfo$avg.width
}

k.values <- 2:10
sil_values <- map_dbl(k.values, sil_width)

plot(k.values, sil_values, type = "b", pch = 19,
     xlab = "Number of clusters",
     ylab = "Average silhouette width")

```

```{r}
set.seed(2025)
kmeans_result <- kmeans(player_clustering, centers = 2, nstart = 25)

player_master_final$Cluster <- kmeans_result$cluster

```

```{r}
run_pca_weights <- function(data_numeric){
  # Perform PCA
  pca <- PCA(data_numeric, graph = FALSE)
  
  # Extract eigenvalues (variance explained by each component)
  eig <- pca$eig
  
  # Extract contributions of variables to PC1 and PC2
  var_contrib <- data.frame(pca$var$contrib)
  
  # Sum contribution of Dim.1 and Dim.2 and scale them to sum to 1
  var_contrib <- var_contrib %>%
    mutate(Total_Contrib = Dim.1 + Dim.2,
           Weight = Total_Contrib / sum(Total_Contrib))
  
  # Return PCA object, weights, and eigenvalues
  return(list(pca = pca, weights = var_contrib, eig = eig))
}

player_clustering_final <- player_clustering
player_clustering_final$Player <- player_master_final$Player
player_clustering_final$Squad <- player_master_final$Squad
player_clustering_final$Comp <- player_master_final$Comp
player_clustering_final$Pos <- player_master_final$Pos
player_clustering_final$Cluster <- player_master_final$Cluster
player_clustering_final$Min_Playing_Time <- player_master_final$Min_Playing_Time


clustered_pca_results <- player_clustering_final %>% 
  group_split(Cluster) %>% 
  map(~run_pca_weights(.x %>% select(DPR_std, OCR_std, GSR_std)))


```

```{r}
cluster_weights <- map(clustered_pca_results, ~.x$weights %>% select(Weight) %>% rownames_to_column("Metric"))
names(cluster_weights) <- paste0("Cluster_", 1:length(cluster_weights))

player_clusters <- player_clustering_final %>% group_split(Cluster)

cluster_ratings <- map2(
  player_clusters,
  cluster_weights,
  ~ .x %>% mutate(
    Overall_Rating =
      DPR_std * .y$Weight[.y$Metric == "DPR_std"] +
      OCR_std * .y$Weight[.y$Metric == "OCR_std"] +
      GSR_std * .y$Weight[.y$Metric == "GSR_std"]
  )
)

player_master_final <- bind_rows(cluster_ratings)
player_master_final <- player_master_final %>% arrange(desc(Overall_Rating))

pl_ratings_final <- player_master_final %>% filter(Comp == "Premier League")
laliga_ratings_final <- player_master_final %>% filter(Comp == "La Liga")
ligue1_ratings_final <- player_master_final %>% filter(Comp == "Ligue 1")
bundesliga_ratings_final <- player_master_final %>% filter(Comp == "Bundesliga")
seriea_ratings_final <- player_master_final %>% filter(Comp == "Serie A")

```

### Finally decide the awards and the team of the year based on minutes - 2260 minutes played (24 games)

```{r}
player_master_awards <- player_master_final %>% filter(Min_Playing_Time >= 2260)
pl_ratings_awards <- pl_ratings_final %>% filter(Min_Playing_Time >= 2260)
laliga_ratings_awards <- laliga_ratings_final %>% filter(Min_Playing_Time >= 2260)
ligue1_ratings_awards <- ligue1_ratings_final %>% filter(Min_Playing_Time >= 2260)
seriea_ratings_awards <- seriea_ratings_final %>% filter(Min_Playing_Time >= 2260)
bundesliga_ratings_awards <- bundesliga_ratings_final %>% filter(Min_Playing_Time >= 2260)

head(player_master_awards)
head(pl_ratings_awards)
head(laliga_ratings_awards)
head(ligue1_ratings_awards)
head(seriea_ratings_awards)
head(bundesliga_ratings_awards)

```

## Average Player Rating of Each League

```{r}
print(median(pl_ratings_final$Overall_Rating))
print(median(laliga_ratings_final$Overall_Rating))
print(median(ligue1_ratings_final$Overall_Rating))
print(median(bundesliga_ratings_final$Overall_Rating))
print(median(seriea_ratings_final$Overall_Rating))


```